EDIpack2: interoperable Lanczos-based solver for generic quantum impurity problems

### 2025-05-06 03_edipack2.tex
GB halfway through the iRDM section (currently 3.10) I found a seemingly wrong statement that I want to discuss with AA.

\subsection{Observables}\label{sSecObc}
A number of predefined impurity observable or local static
correlations, e.g. occupation, total energy, pair amplitude, excitonic
order parameter ,etc.   are calculated in the module
\texttt{OBSERVABLES}. As for the previous cases, this module wraps the
different implementations and quantities to evaluate from the
operational modes {\tt ed\_mode=?} implemented in {\tt
  ED\_OBSERVABLES\_?}. 
Local observables and correlations are defined by the thermal average  $\ibra \OO\iket =
\Tr{\left[e^{-\beta H}\OO\right]}/Z$. This can be efficiently evaluated at zero and
low temperature using the stored low energy part of the spectrum
leveraging on the exponential cutoff from the Boltzmann factor:
$$
\ibra \OO\iket  = \tfrac{1}{Z}\sum_n e^{-\beta
  E_n}\bra{\psi_n}\OO\ket{\psi_n}
$$
where $E_n$ and $\ket{\psi_n}$ are the low lying eigenstates of the
system stored in the {\tt state\_list}.

%GBtodo placeholder: revise this subsection
\subsection{Reduced impurity density matrix}\label{sSecRDM}
This version of the \NAME library features the calculation of the
impurity Reduced Density Matrix (iRDM), extending the algorithm introduced in \onlinecite{BellomiaPhD} so to enable the analysis of
entanglement properties of the quantum impurity for any value of {\tt
  ed\_mode}.
To simplify the discussion in this section we take the zero
temperature limit and consider the presence of a non-degenerate
groundstate $\ket{\psi}$ populating the {\tt state\_list}.
The generalization to the presence of degenerate states or the finite
temperature regime are straightforward.
The groundstate belongs to a given symmetry sector $\SS$ and it
represented in the Fock basis as:  $\ket{\psi} = \sum_I a_I \ket{I}$. 
The density matrix $\rho$ is defined as:
$$
\rho = \ket{\psi}\bra{\psi} = \sum_{nm=1}^{4^{N_\mathrm{s}}} a^*_ma_n \ket{n}\bra{m} =\sum_{nm=1}^{4^{N_\mathrm{s}}} \rho_{nm} \ket{n}\bra{m} 
$$
The iRDM is obtained from this by tracing out the bath degrees of
freedom: $\rho^{i}=Tr_{b}\rho$. The summation appearing in the
previous expressions becomes quickly unmaneagable with the system
size. To overcome this issue we implemented a fast algorithm
leveraging on symmetry sector and map sparsity.
Using quantum number conservation the summation extrema get reducedto
the size $D_S$ of the sector $S$ to which the groundstate (or in
general any eigenstate) belongs. 
Next we analyze the groundstate expansion on the Fock states, splitting the bit representation of any state $\ket{I}$ into
their spin-dependent impurity and bath parts as $\ket{I} = \ket{
  i_\uparrow\,  b_\uparrow\, i_\downarrow \, b_\downarrow}$. Each bit
set span a space of dimension $D^\a_{imp}$ or $D^\s_{bath}$. 
Finally, we store into the {\tt sparse\_map} ${\cal S}$, for every impurity configuration 
$i_\s =1,\dots,D^\s_{imp}$ (key) the corresponding bath states $j\in[1,\dots,D^\s_{bath}]$
(values) which contribute to that key, as detailed in
\secu{CodeSparseMap}.
We indicate with ${\cal D}^\s_{i}$ the number of values for the key $i_\s$.
The fast summation algorithm for the evaluation of the iRDM takes on a
different form for the two cases {\bf normal} and {\bf superc/nonsu2},
as outlined below.

\paragraph{{\bf normal}}
In this case we take advantage of the Fock and sector spin resolved
factorization discussed in \secu{sSecQNs}. As such we can also split
impurity and bath bits independently for each spin configuration:
$$
\ket{\vec{n}} =
\ket{\vec{n}_\up}\otimes \ket{\vec{n}_\dw} =
\ket{\vec{i}_\up}\ket{\vec{b}_\up}\otimes \ket{\vec{i}_\dw}\ket{\vec{b}_\dw}
$$
In addition it should be noted that creation (destruction) operators at
position $p$ for spin $\sigma$  $c^+_{p\sigma}$ ($c^+_{p\sigma}$) do
not operate on the opposite spin so that they pass through without
taking any fermionic sign. Finally, we observe that local iRDM admits
only diagonal terms in the {\bf normal} case. Thus we obtain for the iRDM:
\begin{equation}
  \label{iRDMnormal}
  \begin{aligned}
  \rho^\mathrm{imp} &=
  \sum_{\sigma=\up\dw}\sum_{b_\sigma=1}^{D_{b_\sigma}}
  \bra{b_\up}\otimes\bra{b_\dw}
    \rho
    \ket{b_\dw}\otimes\ket{b_\up}    \cr
    %
    &=\sum_{\sigma=\up\dw}
    \sum_{b_\sigma=1}^{D_{b_\sigma}}
    \sum_{i_\sigma=1}^{D_{i_\sigma}}
    \sum_{p_\sigma=1}^{D_{p_\sigma}}
    \sum_{j_\sigma=1}^{D_{j_\sigma}} 
    \sum_{q_\sigma=1}^{D_{q_\sigma}}
    C_{i,p}\,C_{j,q}\,a_{i_\up p_\up i_\dw p_\dw} a^*_{j_\up q_\up j_\dw q_\dw}
    \langle b_\up| p_\up\rangle| i_\up\rangle\langle j_\up |\langle q_\up| b_\up\rangle   \otimes \langle b_\dw| p_\dw \rangle|i_\dw\rangle \langle j_\dw  |\langle q_\dw| b_\dw\rangle
    \cr
&=
    \sum_{b_\sigma=1}^{D_{b_\sigma}}
    \sum_{i_\sigma=1}^{D_{i_\sigma}}
    a_{i_\up b_\up i_\dw b_\dw} a^*_{i_\up b_\up i_\dw b_\dw}
    \ket{i_\up}\bra{i_\up}\otimes \ket{i_\dw}\bra{i_\dw}
  \end{aligned}
\end{equation}
where the constant $C_{i,b}$ take care of the sign change due to
exchanging bath and impurity bitsets. As effect of the Kronecker
delta symbols for bath and impurity these constants are identically 1.  
The numerical implementation relies on the use of the {\tt
  sparse\_map} ($\%sp$) for the groundstate sector as reported in the
listing below:  
\begin{lstlisting}[style=fstyle,numbers=none,basicstyle={\scriptsize\ttfamily}]
do IimpUp=0,2**Norb-1
  do JimpUp=0,2**Norb-1
    !Finding the unique bath states connecting IimpUp and JimpUp
    call sp_return_intersection(sectorI%H(1)%sp,IimpUp,JimpUp,BATHup,lenBATHup)
    if(lenBATHup==0)cycle
    do IimpDw=0,2**Norb-1
      do JimpDw=0,2**Norb-1
        !Finding the unique bath states connecting IimpDw and JimpDw -> BATHdw(:)
        call sp_return_intersection(sectorI%H(2)%sp,IimpDw,JimpDw,BATHdw,lenBATHdw)
        if(lenBATHdw==0)cycle
        do ibUP=1,lenBATHup
          IbathUp = BATHup(ibUP)
          do ibDW=1,lenBATHdw
            IbathDw = BATHdw(ibDW)
            !Allowed spin Fock space Istates:
            !Iup = IimpUp +  2^Norb * IbathUp
            !Idw = IimpDw +  2^Norb * IbathDw
            iUP= binary_search(sectorI%H(1)%map,IimpUp + 2**Norb*IbathUp)
            iDW= binary_search(sectorI%H(2)%map,IimpDw + 2**Norb*IbathDw)
            i  = iUP + (iDW-1)*sectorI%DimUp
            !Allowed spin Fock space Jstates:
            !Jup = JimpUp +  2^Norb * IbathUp
            !Jdw = JimpDw +  2^Norb * IbathDw
            jUP= binary_search(sectorI%H(1)%map,JimpUp + 2**Norb*IbathUp)
            jDW= binary_search(sectorI%H(2)%map,JimpDw + 2**Norb*IbathDw)
            j  = jUP + (jDW-1)*sectorI%DimUp
            ! 
            io = (IimpUp + 2**Norb*IimpDw) + 1
            jo = (JimpUp + 2**Norb*JimpDw) + 1
            irdm(io,jo) = irdm (io,jo) + psi(i)*psi(j)*weight
          enddo
        enddo
      enddo
    enddo
  enddo
enddo
\end{lstlisting}

@@

ordering of the key-value pairs in the {\tt sparse\_map} in this case
follow a contiguous ordering and that their decomposition is necessary
to reconstruct the correct Fock state index, see the following

>>>>

\subsection{Observables}\label{sSecObc}
A number of predefined impurity observable or local static
correlations, e.g. occupation, total energy, pair amplitude, excitonic
order parameter ,etc.   are calculated in the module
\texttt{OBSERVABLES}. As for the previous cases, this module wraps the
different implementations and quantities to evaluate from the
operational modes {\tt ed\_mode=?} implemented in {\tt
  ED\_OBSERVABLES\_?}. 
Local observables and correlations are defined by the thermal average  $\ibra \OO\iket =
\Tr{\left[e^{-\beta H}\OO\right]}/Z$. This can be efficiently evaluated at zero and
low temperature using the stored low energy part of the spectrum
leveraging on the exponential cutoff from the Boltzmann factor:
\begin{equation}
    \ibra \OO\iket  = \tfrac{1}{Z}\sum_n e^{-\beta
  E_n}\bra{\psi_n}\OO\ket{\psi_n},
  \label{eq:thermal_average}
\end{equation}
where $E_n$ and $\ket{\psi_n}$ are the low lying eigenstates of the
system stored in the {\tt state\_list}.

\subsection{Reduced impurity density matrix}\label{sSecRDM}
This version of the \NAME library features the calculation of the
impurity Reduced Density Matrix (iRDM, $\rho^\mathrm{imp}$), extending 
the algorithm introduced in \onlinecite{BellomiaPhD} so to enable the
analysis of entanglement properties of the quantum impurity for any value
of {\tt ed\_mode}.
To simplify the discussion in this section we take the zero
temperature limit and consider the presence of a non-degenerate
groundstate $\ket{\psi}$ populating the {\tt state\_list}.
The generalization to the presence of degenerate states or the finite
temperature regime are straightforward, as it amounts to defining an
average over degenerate/thermal states $\ibra\rho^\mathrm{imp}\iket$, 
akin to \equ{eq:thermal_average}.
The given pure state belongs to a unique symmetry sector $\SS$ and it
is represented in the Fock basis as $\ket{\psi} = \sum_I a_I \ket{I}$. 
The corresponding pure density matrix $\rho$ is thus explicitly decomposed as:
$$
\rho = \ket{\psi}\bra{\psi} = \sum_{nm=1}^{4^{N_\mathrm{s}}} a^*_ma_n \ket{n}\bra{m} =\sum_{nm=1}^{4^{N_\mathrm{s}}} \rho_{nm} \ket{n}\bra{m}.
$$
The iRDM is obtained from $\rho$ by tracing out the bath degrees of
freedom: $\rho^\mathrm{imp}=\Tr_\mathrm{bath}\rho$. The summations appearing in the
previous expressions become quickly unmanageable with the system
size, as their extrema scale exponentially with $N_\mathrm{s}$. 
To overcome this issue we implemented a fast algorithm
leveraging on the decomposition of the Fock space into symmetry sectors
and the subsequent map sparsity.
Using quantum number conservation the summation extrema get reduced to
the size $D_\SS$ of the sector $\SS$ to which the groundstate (or in
general any eigenstate) belongs. 
Next we analyze the groundstate expansion on the Fock states, splitting the bit representation of any state $\ket{I}$ into
their spin-dependent impurity and bath parts as $\ket{I} = \ket{
  i_\uparrow\,  b_\uparrow\, i_\downarrow \, b_\downarrow}$. Each bitset 
span a space of dimension $D^\a_{imp}$ or $D^\s_{bath}$. 
Finally, as detailed in \secu{CodeSparseMap}, we store into the {\tt sparse\_map} ${\cal S}$, 
for every impurity configuration 
$i_\s =1,\dots,D^\s_{imp}$ (\emph{key}), the bath states 
$j\in[1,\dots,D^\s_{bath}]$ (\emph{values}) that satisfy the conservation law defined
by the \texttt{ed\_mode} variable.
We indicate with ${\cal D}^\s_{i}$ the number of values for the key $i_\s$.
The fast summation algorithm for the evaluation of the iRDM takes on a
different form for the two cases {\bf normal} and {\bf superc/nonsu2},
as outlined below.

%GBtodo placeholder: revise this subsection from this point onwards
\paragraph{{\bf normal}}
In this case we take advantage of the Fock space and sector spin-resolved
factorization, discussed in depth in \secu{sSecQNs}. We can indeed split the
global bitset in spin-resolved impurity and bath states, as
$$
\ket{\vec{n}} =
\ket{\vec{n}_\up}\otimes \ket{\vec{n}_\dw} =
\ket{\vec{i}_\up}\ket{\vec{b}_\up}\otimes \ket{\vec{i}_\dw}\ket{\vec{b}_\dw}.
$$
We note that creation (destruction) operators at
position $p$ for spin $\sigma$  $c^+_{p\sigma}$ ($c^+_{p\sigma}$) do
not operate on the opposite spin $\bar{\sigma}$ so that they pass through 
the $\ket{n_{\bar{\sigma}}}$ bitset without taking any fermionic sign. 
Finally, we observe that local iRDM admits
only {\color{red} diagonal} terms in the {\bf normal} case. {\color{red} [GB: this is a wrong statement: do you mean \emph{spin-}diagonal?]}

Thus we obtain for the iRDM:
\begin{equation}
  \label{iRDMnormal}
  \begin{aligned}
  \rho^\mathrm{imp} &=
  \sum_{\sigma=\up\dw}\sum_{b_\sigma=1}^{D_{b_\sigma}}
  \bra{b_\up}\otimes\bra{b_\dw}
    \rho
    \ket{b_\dw}\otimes\ket{b_\up}    \cr
    %
    &=\sum_{\sigma=\up\dw}
    \sum_{b_\sigma=1}^{D_{b_\sigma}}
    \sum_{i_\sigma=1}^{D_{i_\sigma}}
    \sum_{p_\sigma=1}^{D_{p_\sigma}}
    \sum_{j_\sigma=1}^{D_{j_\sigma}} 
    \sum_{q_\sigma=1}^{D_{q_\sigma}}
    C_{i,p}\,C_{j,q}\,a_{i_\up p_\up i_\dw p_\dw} a^*_{j_\up q_\up j_\dw q_\dw}
    \langle b_\up| p_\up\rangle| i_\up\rangle\langle j_\up |\langle q_\up| b_\up\rangle   \otimes \langle b_\dw| p_\dw \rangle|i_\dw\rangle \langle j_\dw  |\langle q_\dw| b_\dw\rangle
    \cr
&=
    \sum_{b_\sigma=1}^{D_{b_\sigma}}
    \sum_{i_\sigma=1}^{D_{i_\sigma}}
    a_{i_\up b_\up i_\dw b_\dw} a^*_{i_\up b_\up i_\dw b_\dw}
    \ket{i_\up}\bra{i_\up}\otimes \ket{i_\dw}\bra{i_\dw}
  \end{aligned}
\end{equation}
where the constant $C_{i,b}$ take care of the sign change due to
exchanging bath and impurity bitsets. As effect of the Kronecker
delta symbols for bath and impurity these constants are identically 1.  
The numerical implementation relies on the use of the {\tt
  sparse\_map} ($\%sp$) for the groundstate sector as reported in the
listing below:  
\begin{lstlisting}[style=fstyle,numbers=none,basicstyle={\scriptsize\ttfamily}]
do IimpUp=0,2**Norb-1
  do JimpUp=0,2**Norb-1
    !Finding the unique bath states connecting IimpUp and JimpUp
    call sp_return_intersection(sectorI%H(1)%sp,IimpUp,JimpUp,BATHup,lenBATHup)
    if(lenBATHup==0)cycle
    do IimpDw=0,2**Norb-1
      do JimpDw=0,2**Norb-1
        !Finding the unique bath states connecting IimpDw and JimpDw -> BATHdw(:)
        call sp_return_intersection(sectorI%H(2)%sp,IimpDw,JimpDw,BATHdw,lenBATHdw)
        if(lenBATHdw==0)cycle
        do ibUP=1,lenBATHup
          IbathUp = BATHup(ibUP)
          do ibDW=1,lenBATHdw
            IbathDw = BATHdw(ibDW)
            !Allowed spin Fock space Istates:
            !Iup = IimpUp +  2^Norb * IbathUp
            !Idw = IimpDw +  2^Norb * IbathDw
            iUP= binary_search(sectorI%H(1)%map,IimpUp + 2**Norb*IbathUp)
            iDW= binary_search(sectorI%H(2)%map,IimpDw + 2**Norb*IbathDw)
            i  = iUP + (iDW-1)*sectorI%DimUp
            !Allowed spin Fock space Jstates:
            !Jup = JimpUp +  2^Norb * IbathUp
            !Jdw = JimpDw +  2^Norb * IbathDw
            jUP= binary_search(sectorI%H(1)%map,JimpUp + 2**Norb*IbathUp)
            jDW= binary_search(sectorI%H(2)%map,JimpDw + 2**Norb*IbathDw)
            j  = jUP + (jDW-1)*sectorI%DimUp
            ! 
            io = (IimpUp + 2**Norb*IimpDw) + 1
            jo = (JimpUp + 2**Norb*JimpDw) + 1
            irdm(io,jo) = irdm (io,jo) + psi(i)*psi(j)*weight
          enddo
        enddo
      enddo
    enddo
  enddo
enddo
\end{lstlisting}

@@

ordering of the \emph{key-value} pairs in the {\tt sparse\_map} in this case
follows a contiguous ordering and that their decomposition is necessary
to reconstruct the correct Fock state index. See the following
